---
title: "Forests and Water"
author: "Aaron Berdanier, Nicholas School of Environment"
date: "11/20/2015"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(raster)
library(rgdal)
library(sp)
load("GISday.Rdata")
```

# GIS and R

## A GIS is powerful. R is powerful. Together they are a force to be reckoned with.

***

# An example with USFS Forest Inventory and Analysis

## Get all of the FIA plots in the United States from the internet.

```{r eval=FALSE}
library(raster)
library(rgdal)
library(sp)
plotdat <- matrix(NA,nrow=0,ncol=57)
states <- state.abb[which(!state.abb%in%c("HI","AK"))]
prg <- txtProgressBar(min = 0, max = length(states), char = "+",style = 3)
for(s in 1:length(states)){
  st <- states[s]
  url <- paste0("http://apps.fs.fed.us/fiadb-downloads/",st,"_PLOT.csv")
  tmp <- read.csv(url)
  plotdat <- rbind(plotdat,tmp)
  setTxtProgressBar(prg, s)  
}
close(prg)

# only forested plots with coordinates
forestdat <- plotdat[which(plotdat$PLOT_STATUS_CD==1&!is.na(plotdat$LON)),] 
```

## Where are our forests?

```{r}
# number of plots
nrow(forestdat) 

# projected coordinates, longlat, datum & ellps WGS84
coo <- SpatialPoints(forestdat[,c("LON","LAT")],CRS("+init=epsg:4326"))
par(mar=c(0,0,0,0))
plot(coo,pch=".",col="#00000010")
```


# Predicting forests
###  **Question:** Drought impacts on forest extent?

#### **Data from** Dobrowski, S.Z., J. Abatzoglou, A.K. Swanson, J.A. Greenberg, A.R. Mynsberge, Z.A. Holden, M.K. Schwartz (2013) The climate velocity of the contiguous United States during the 20th century. Global Change Biology 19: 241-251. [Download from internet](http://adaptwest.databasin.org/pages/adaptwest-waterbalance)

```{r eval=FALSE}
# load data
ff <- list.files("~/Downloads/def_2000_2009/")
def <- stack(paste0("~/Downloads/def_2000_2009/",ff))
# average across years
mndef <- stackApply(def,rep(1,nlayers(def)),fun="mean")

# reproject to match forest plots
defpr <- projectRaster(mndef,crs=projection(coo))
defpro <- aggregate(defpr,12) # to 0.1 degrees
```

```{r echo=FALSE}
par(mar=c(0,0,2,0))
plot(defpro, main="Climatic water deficit (PET - AET)", col=terrain.colors(255))
```

## Match FIA plots to climate map by *rasterizing* 
#### Sort of like this:

```{r echo=FALSE}
ext <- extent(defpro)
par(mar=c(0,0,0,0))
plot(coo,pch=".",col="#00000010")
abline(v=seq(ext[1],ext[2],length.out=dim(defpro)[2]/10),col="#00000066")
abline(h=seq(ext[3],ext[4],length.out=dim(defpro)[1]/10),col="#00000066")
#
```

```{r}
cooras <- rasterize(coo,defpro,
                    fun=function(x,...) if(length(x)>0) 1, na.rm=T)
cooras[is.na(cooras)] <- 0
cooras[is.na(values(defpro))] <- NA

```

## Then, a statistical model...

```{r}
forestmat <- data.frame(deficit=values(defpro), forest=values(cooras))
fmod <- glm(forest~deficit, data=forestmat, family=binomial())
summary(fmod)
```

```{r echo=FALSE}
par(mar=c(4,4,0.4,0.4))
boxplot(deficit~forest, data=forestmat, at=c(0,1),
        horizontal=T,las=1,staplewex=0,lty=1,outline=F,frame=F,
        ylab="Forest?",xlab="Climatic water deficit (PET - AET)")
defpre <- seq(1,2500)
forpre <- predict(fmod,data.frame(deficit=defpre),type="response")
lines(forpre~defpre,lwd=3)
forval <- -coef(fmod)[1]/coef(fmod)[2]
abline(v=forval, lwd=5, col="#001A57")
```

## Now, we can make some predictions

```{r}
# transformed to US National Atlas Equal Area
coopr <- spTransform(coo,CRS("+init=epsg:2163"))
defplot <- projectRaster(defpro,crs="+init=epsg:2163")
```

```{r echo=FALSE}
par(mar=c(0,0,2,0))
plot(defplot, col=terrain.colors(255),main="Climatic water deficit (PET - AET)")
```

```{r echo=FALSE}
par(mar=c(0,0,2,0))
plot(defplot, col=terrain.colors(255),main="Deficit + FIA plots")
points(coopr,pch=".",col="#00000015")
```

```{r echo=FALSE}
par(mar=c(0,0,2,0))
plot(defplot, col=terrain.colors(255),main="Deficit + Plots + Prediction")
points(coopr,pch=".",col="#00000015")
contour(defplot,nlevels=1,labels="Pr=0.5",levels=forval,add=T,lwd=3,col="#001A57")
```

## Feel free to email me: abb30@duke.edu